---
title: 'Assignment 4: Text and Sequence Data'
author: "Brian Yim"
date: "2026-04-18"
output:
  word_document: default
  html_document: default
---


```{r}
library(keras3)
library(tensorflow)
library(reticulate)
library(tidyverse)
```
#loading the data
```{r}
max_features <- 10000
maxlen <- 150

imdb <- dataset_imdb(num_words = max_features)

c(c(x_train, y_train), c(x_test, y_test)) %<-% imdb
```
#restricting training data and validation data
```{r}
set.seed(123)

x_val <- x_train[1:10000]
partial_x_train <- x_train[10001:(10000 + 100)]

y_val <- y_train[1:10000]
partial_y_train <- y_train[10001:(10000 + 100)]
```

#Pad sequences so every review has length 150
```{r}
x_val <- pad_sequences(x_val, maxlen = maxlen)
partial_x_train <- pad_sequences(partial_x_train, maxlen = maxlen)
x_test <- pad_sequences(x_test, maxlen = maxlen)
```

#now I build model 1 which is regular embedding model
```{r}
build_embedding_model <- function() {
  model <- keras_model_sequential() |>
    layer_embedding(input_dim = max_features, output_dim = 16, input_length = maxlen) |>
    layer_flatten() |>
    layer_dense(units = 16, activation = "relu") |>
    layer_dense(units = 1, activation = "sigmoid")

  model |> compile(
    optimizer = "rmsprop",
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

  model
}
```

#train this model
```{r}
model_embed <- build_embedding_model()

history_embed <- model_embed |> fit(
  partial_x_train,
  partial_y_train,
  epochs = 20,
  batch_size = 32,
  validation_data = list(x_val, y_val),
  verbose = 2
)
```

```{r}
plot(history_embed)
```

#I downloaded pretrained word vector glove.6B.100d. Then I got the IMDB word index
```{r}
word_index <- dataset_imdb_word_index()
word_index <- lapply(word_index, function(x) x + 3)
word_index["[PAD]"] <- 0
word_index["[START]"] <- 1
word_index["[UNK]"] <- 2
word_index["[UNUSED]"] <- 3
```

#read the glove.6B.100d.txt
```{r}
embeddings_index <- new.env(hash = TRUE)

lines <- readLines("glove.6B.100d.txt")

for (line in lines) {
  values <- strsplit(line, " ")[[1]]
  word <- values[1]
  coefs <- as.numeric(values[-1])
  embeddings_index[[word]] <- coefs
}
```

#created the matrix
```{r}
embedding_dim <- 100
embedding_matrix <- matrix(0, nrow = max_features, ncol = embedding_dim)

for (word in names(word_index)) {
  index <- word_index[[word]]
  if (index < max_features) {
    embedding_vector <- embeddings_index[[word]]
    if (!is.null(embedding_vector)) {
      embedding_matrix[index + 1, ] <- embedding_vector
    }
  }
}
```


#Then, the pretrained embedding model was built
```{r}
build_pretrained_model <- function(embedding_matrix) {
  model <- keras_model_sequential() |>
    layer_embedding(
      input_dim = max_features,
      output_dim = embedding_dim,
      input_length = maxlen,
      weights = list(embedding_matrix),
      trainable = FALSE
    ) |>
    layer_flatten() |>
    layer_dense(units = 16, activation = "relu") |>
    layer_dense(units = 1, activation = "sigmoid")

  model |> compile(
    optimizer = "rmsprop",
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

  model
}
```

#trained this model
```{r}
model_pretrained <- build_pretrained_model(embedding_matrix)

history_pretrained <- model_pretrained |> fit(
  partial_x_train,
  partial_y_train,
  epochs = 20,
  batch_size = 32,
  validation_data = list(x_val, y_val),
  verbose = 2
)
```

#then, compared the two models
```{r}
best_embed <- max(history_embed$metrics$val_accuracy)
best_pretrained <- max(history_pretrained$metrics$val_accuracy)

results_100 <- data.frame(
  training_samples = 100,
  model = c("Learned Embedding", "Pretrained Embedding"),
  best_val_accuracy = c(best_embed, best_pretrained)
)

results_100
```

#Next, I will compare different training size (100, 500, 1000, 2500, 5000)
```{r}
run_experiment <- function(n_train, x_train, y_train, x_val, y_val, embedding_matrix) {
  
  partial_x_train <- x_train[10001:(10000 + n_train)]
  partial_y_train <- y_train[10001:(10000 + n_train)]
  
  partial_x_train <- pad_sequences(partial_x_train, maxlen = maxlen)
  
  # Learned embedding model
  model_embed <- build_embedding_model()
  history_embed <- model_embed |> fit(
    partial_x_train,
    partial_y_train,
    epochs = 20,
    batch_size = 32,
    validation_data = list(x_val, y_val),
    verbose = 0
  )
  
  best_embed <- max(history_embed$metrics$val_accuracy)
  
  # Pretrained embedding model
  model_pretrained <- build_pretrained_model(embedding_matrix)
  history_pretrained <- model_pretrained |> fit(
    partial_x_train,
    partial_y_train,
    epochs = 20,
    batch_size = 32,
    validation_data = list(x_val, y_val),
    verbose = 0
  )
  
  best_pretrained <- max(history_pretrained$metrics$val_accuracy)
  
  data.frame(
    training_samples = n_train,
    learned_embedding = best_embed,
    pretrained_embedding = best_pretrained
  )
}
```

```{r}
sample_sizes <- c(100, 500, 1000, 2500, 5000)

all_results <- lapply(sample_sizes, run_experiment,
                      x_train = x_train,
                      y_train = y_train,
                      x_val = x_val,
                      y_val = y_val,
                      embedding_matrix = embedding_matrix)

all_results <- bind_rows(all_results)
all_results
```

```{r}
results_long <- all_results |>
  pivot_longer(cols = c(learned_embedding, pretrained_embedding),
               names_to = "model",
               values_to = "best_val_accuracy")

ggplot(results_long, aes(x = training_samples, y = best_val_accuracy, color = model)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Validation Accuracy by Training Sample Size",
    x = "Training Sample Size",
    y = "Best Validation Accuracy"
  ) +
  theme_minimal()
```

