---
title: "Assignment 3"
author: "Brian Yim"
date: "2026-04-05"
output:
  html_document: default
  word_document: default
---

#For this timeseries assignment I used jena_climate_2009_2016.csv dataset. 




```{r}
library(keras3)
library(tensorflow)
library(tidyverse)
```


```{r}
# Load data
data <- read.csv("C:/Users/hyim/OneDrive - Kent State University/Desktop/Advanced Machine Learning/jena_climate_2009_2016.csv")
head(data)
str(data)
```
#Next, I converted the data
```{r}
data$Date.Time <- as.POSIXct(data$Date.Time, format = "%d.%m.%Y %H:%M:%S")
```

#Next, I selected 4 items "temp", "humidity", "wind", "pressure" for the assignment
```{r}
data_numeric <- data %>%
  select(
    T..degC.,     # temperature (target)
    rh....,       # humidity
    wv..m.s.,     # wind speed
    p..mbar.      # pressure
  )
```

#And then, renamed them
```{r}
colnames(data_numeric) <- c("temp", "humidity", "wind", "pressure")
```

#Next, I split them into tain (50%), validate (25%), and test (25%) dataset
```{r}
n <- nrow(data_numeric)

train_end <- floor(0.5 * n)
val_end   <- floor(0.75 * n)

train_data <- data_numeric[1:train_end, ]
val_data   <- data_numeric[(train_end + 1):val_end, ]
test_data  <- data_numeric[(val_end + 1):n, ]
```

#to move forward, I normalized them
```{r}
train_mean <- apply(train_data, 2, mean)
train_sd   <- apply(train_data, 2, sd)

train_scaled <- scale(train_data, center = train_mean, scale = train_sd)
val_scaled   <- scale(val_data, center = train_mean, scale = train_sd)
test_scaled  <- scale(test_data, center = train_mean, scale = train_sd)
```

#The current timeseries in in 10 minutes which is quite frequent hence too dense. So I selected every 6 rows meaning I am analyzing every hour
```{r}
sampling_rate <- 6

train_scaled <- train_scaled[seq(1, nrow(train_scaled), by = sampling_rate), ]
val_scaled   <- val_scaled[seq(1, nrow(val_scaled), by = sampling_rate), ]
test_scaled  <- test_scaled[seq(1, nrow(test_scaled), by = sampling_rate), ]
```

#Next, I created the sequence. Past 120 time steps (~5 days) to predict next temp
```{r}
sequence_length <- 120
target_col <- 1
batch_size <- 32
```

#Then created the dataset
```{r}
train_ds <- timeseries_dataset_from_array(
  data = train_scaled[1:(nrow(train_scaled) - 1), ],
  targets = train_scaled[(sequence_length + 1):nrow(train_scaled), target_col],
  sequence_length = sequence_length,
  batch_size = batch_size
)

val_ds <- timeseries_dataset_from_array(
  data = val_scaled[1:(nrow(val_scaled) - 1), ],
  targets = val_scaled[(sequence_length + 1):nrow(val_scaled), target_col],
  sequence_length = sequence_length,
  batch_size = batch_size
)

test_ds <- timeseries_dataset_from_array(
  data = test_scaled[1:(nrow(test_scaled) - 1), ],
  targets = test_scaled[(sequence_length + 1):nrow(test_scaled), target_col],
  sequence_length = sequence_length,
  batch_size = batch_size
)
```
#GRU baseline Model
```{r}
num_features <- ncol(train_scaled)

model_gru <- keras_model_sequential() %>%
  layer_gru(
    units = 32,
    input_shape = c(sequence_length, num_features),
    dropout = 0.2,
    recurrent_dropout = 0.2
  ) %>%
  layer_dense(units = 1)

model_gru %>% compile(
  optimizer = "rmsprop",
  loss = "mse",
  metrics = c("mae")
)

history_gru <- model_gru %>% fit(
  train_ds,
  validation_data = val_ds,
  epochs = 20
)
```

```{r}
callbacks_list <- list(
  callback_early_stopping(
    monitor = "val_mae",
    patience = 5,
    restore_best_weights = TRUE
  )
)
```


#Next model is stacked GRU model
```{r}
model_stacked_gru <- keras_model_sequential() %>%
  layer_gru(
    units = 32,
    return_sequences = TRUE,
    input_shape = c(sequence_length, num_features),
    dropout = 0.2,
    recurrent_dropout = 0.2
  ) %>%
  layer_gru(
    units = 32,
    dropout = 0.2,
    recurrent_dropout = 0.2
  ) %>%
  layer_dense(units = 1)

model_stacked_gru %>% compile(
  optimizer = "rmsprop",
  loss = "mse",
  metrics = c("mae")
)

history_stacked_gru <- model_stacked_gru %>% fit(
  train_ds,
  validation_data = val_ds,
  epochs = 20,
  callbacks = callbacks_list
)
```
#Next, I ran LSTM
```{r}
model_lstm <- keras_model_sequential() %>%
  layer_lstm(
    units = 32,
    input_shape = c(sequence_length, num_features),
    dropout = 0.2,
    recurrent_dropout = 0.2
  ) %>%
  layer_dense(units = 1)

model_lstm %>% compile(
  optimizer = "rmsprop",
  loss = "mse",
  metrics = c("mae")
)

history_lstm <- model_lstm %>% fit(
  train_ds,
  validation_data = val_ds,
  epochs = 20,
  callbacks = callbacks_list
)
```
#Conv1D + GRU Model
```{r}
model_conv_gru <- keras_model_sequential() %>%
  layer_conv_1d(
    filters = 32,
    kernel_size = 5,
    activation = "relu",
    input_shape = c(sequence_length, num_features)
  ) %>%
  layer_max_pooling_1d(pool_size = 2) %>%
  layer_gru(
    units = 32,
    dropout = 0.2,
    recurrent_dropout = 0.2
  ) %>%
  layer_dense(units = 1)

model_conv_gru %>% compile(
  optimizer = "rmsprop",
  loss = "mse",
  metrics = c("mae")
)

history_conv_gru <- model_conv_gru %>% fit(
  train_ds,
  validation_data = val_ds,
  epochs = 20,
  callbacks = callbacks_list
)
```

#Next, I compared the model. Extract the best validation MAE for each model.
```{r}
get_best_val_mae <- function(history_object) {
  min(as.numeric(history_object$metrics$val_mae))
}
```

#Create a table
```{r}
results <- tibble(
  Model = c("Baseline GRU", "Stacked GRU", "LSTM", "Conv1D + GRU"),
  Best_Val_MAE = c(
    get_best_val_mae(history_gru),
    get_best_val_mae(history_stacked_gru),
    get_best_val_mae(history_lstm),
    get_best_val_mae(history_conv_gru)
  )
) %>%
  arrange(Best_Val_MAE)

results
```

#Finding the best model
```{r}
best_model_name <- results$Model[1]
best_model_name
```
#Next, assign the actual model object
```{r}
best_model <- switch(
  best_model_name,
  "Baseline GRU" = model_gru,
  "Stacked GRU" = model_stacked_gru,
  "LSTM" = model_lstm,
  "Conv1D + GRU" = model_conv_gru
)
```


#Evaluate the best model on the test set
```{r}
test_results <- best_model %>% evaluate(test_ds)
test_results
```
#Add test MAE to the results table
```{r}
results$Test_MAE <- NA
results$Test_MAE[results$Model == best_model_name] <- as.numeric(test_results["mae"])

results
```

#Create a graph
```{r}
ggplot(results, aes(x = reorder(Model, Best_Val_MAE), y = Best_Val_MAE)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Best Validation MAE by Model",
    x = "Model",
    y = "Best Validation MAE"
  ) +
  theme_minimal()
```
#save it
```{r}
ggsave("assignment3_model_comparison.png", width = 8, height = 5)
```

#plot training history
```{r}
plot(history_gru)
plot(history_stacked_gru)
plot(history_lstm)
plot(history_conv_gru)
```

#save the results table
```{r}
write.csv(results, "assignment3_results.csv", row.names = FALSE)
```


#Convert MAE back
```{r}
temp_sd <- train_sd[1]
final_test_mae_original_units <- 0.1507215 * temp_sd
final_test_mae_original_units
```

